Effect of Black Hole Spins on Christodoulou Memory in Binary Mergers

### 2019-04-18 notes/Paper.tex
Ashok made some corrections in the paper

\stm{note everywhere I have added ``the'', that ``upto'' should be ``up to'', etc}
\begin{abstract}
	The Christodoulou memory, which is a nonlinear memory effect sourced by the oscillatory component of the gravitational wave stress tensor, produces a growing, nonoscillatory change in the gravitational wave strain. This results in the permanent displacement of a pair of freely falling test masses after the gravitational wave has passed. While the memory contribution during the early inspiral is well described by the post-Newtonian (PN) approximation, the contribution at merger must be obtained using numerical relativity simulations. The corrections up to third PN order (i.~e.~3PN) for the gravitational wave memory from pairs of nonspinning black holes on quasicircular orbits has been computed by Favata \cite{Favata2009}. Here we include the contribution from the individual black-hole spins up to 1.5PN \stm{what is leading order here, if we then say the memory is 3PN?} and calculate the memory effect up to 3PN order. We also compute the memory contribution during
the merger by using the methods presented in \cite{Favata2009} to calculate the memory content using the oscillatory radiation as input, and
applying that method using publicly available numerical relativity simulations of spinning binaries.
\end{abstract}

@@

where $G^{2-20}_{l'm'lm'-m''-m}$ is the angular integral of equation \ref{eq:2.9} and is given in Appendix \ref{Appendix A}. The above equation can be used to compute the first time derivative of mass-multipole moment by substituting the GW modes $h_{lm}$. When evaluating equation \ref{eq:2.10}, we are only interested in $m=0$ modes. The $m\neq0$ terms yield oscillatory contribution to the waveform polarizations that enter at higher PN orders then the nonoscillatory, $m=0$ terms. The non-vanishing, non-oscillatory modes are $U^{(mem)}_{l0}$ with l-even. These $U^{(mem)}_{l0}$ consist of time-integral of polynomials in $x\equiv(M\omega)^{2/3}$ (where $\omega$ is the orbital angular frequency), with each terms of the form
\begin{equation}\label{eq:2.11}
\int_{-\infty}^{T_{R}} x^{n}dt=\int_{-\infty}^{T_{R}}\frac{x^{n}}{\dot{x}}dx
\end{equation}
After performing the change of variable indicated in above equation \ref{eq:2.11}, evaluation of the resulting integrals requires a model for the frequency evolution of the binary. The adiabatic evolution of the frequency (or $x$) is easily derived from energy balance ($\mathcal{L}_{GW}=-\dot{E}$). and the relation $\dot{x}=-\mathcal{L}_{GW}/(dE/dx)$. The 3.5PN orbital energy and flux are given in Appendix C of reference \cite{Arun2009}. Here we who only up to 1.5 PN terms, the complete expression are show in Appendix \ref{Appendix:C1}. The 1.5 PN expression for energy is given by 
\begin{widetext}
	\begin{align}\label{eq:E}
	E&=-\frac{\eta \, M \, x}{2}\left\{1 + x \left[-\frac{3}{4}-\frac{\eta}{12}\right] + x^{3/2}\left[\left(\frac{8}{3} - \frac{4 \, \eta}{3}\right)\chisdL + \frac{8}{3}\delta\chiadL\right]\right\} \nonumber \\
	\end{align}

@@

\begin{align}\label{eq:2.14}
\frac{dx}{dt} & = \frac{64\, x^5  \, \eta}{5} \left\{1 + x \left[-\frac{743}{336}-\frac{11 \eta }{4} \right] 
+ x^{3/2} \left[4 \pi-\frac{113}{12}\delta \, \chiadL- \left(-\frac{113}{19} + \frac{19 \, \eta}{3}  \right)\chisdL \right] \right\} \nonumber \\ 
\end{align}
\end{widetext}
The results up to 3.5 PN are shows in Appendix \ref{Appendix:C}.
The equation \ref{eq:2.14} can be used to change the integration with respect to time into integration with respect to $x$ as shown in equation in \ref{eq:2.11}. The expressions for $h_{lm}$ whose first time derivatives appear equation \ref{eq:2.10} can be obtained from Ref\cite{Arun2009}. The GW modes for precessing binaries with small inclination angle on nearly circular orbits through 1.5 PN order are given in section IV of ref \cite{Arun2009}. Setting $\iota = 0, \alpha = \pi$ in equation 4.16 and 4.17 gives the result for non-precessing case. In that case the spins are aligned along $z$ axis, so we set $\chi_a^x=\chi_s^x=\chi_a^y=\chi_s^y=0$. We can now write the expression as 
\begin{equation}
h_{lm} = (-1)^{m+1}\frac{(2M\eta v^2)}{D_L}\sqrt{\frac{16\pi}{5}}e^{\mathit{i}\,m\,\psi}\hat{h}_{lm},
\end{equation}
and $\hat{h}_{lm}$ modes are given by 
\begin{subequations}
	\begin{align}\label{eq:2.15}
	\hat{h}_{22} &= 1 + x\left(-\frac{107}{42} + \frac{55\eta}{42}\right) + x^{3/2}\left(2\pi - \frac{4\,\delta\,\chi^{z}_{a}}{3} + \frac{4}{3}\left(-1+\eta\right)\chi^z_s\right)\\
	\hat{h}_{21} &= \frac{\mathit{i}\delta}{3}\left[x^{1/2} - x\frac{3}{2\delta}\left(\chi^{z}_{a} + \delta\chi^{z}_{s}\right) + x^{3/2}\left(-\frac{17}{28} + \frac{5\eta}{7}\right)\right]\\
	\hat{h}_{32} &= \frac{9}{8}\sqrt{\frac{5}{7}}\left[\frac{8}{27}x\left(1-3\nu\right) + \frac{32}{27}x^{3/2}\eta\,\chi^z_s\right]
	\end{align}
\end{subequations}
The remaining $h_{lm}$ expressions does not contain spin terms and the complete expressions up to 3PN terms can be directly taken from Ref.\cite{Blanchet_Iyer2008}
 We show the $h_{lm}$ expressions for positive $m$ values, but the results for negative $m$ values can be obtained by the relation, $h_{l, -m} = (-1)^l \bar{h}_{lm}$. This results is not in general true for precessing binaries, but it continues to hold true of non-precessing case. The time derivative of $h_{lm}$ modes can be obtained by following the procedure shown in Appendix \ref{Appendix B}. With these modes we use equation \ref{eq:2.10} to evaluate the first time derivative of the memory mass-multipole moment, which is further integrated using equation \ref{eq:2.11}. The results after integration are show in next section.

@@

The above equation requires a time integration of absolute value of $\dot{h}_{22}$, but this is not typically the quantity which is directly computed in numerical simulation. The output of NR simulation is usually in the form of component of curvature tensor, or Zerilli-Moncrief-type variables defined relative to a background. 
\subsection{Evaluating Bondi News from numerical data}
The Bondi News which also the first derivative of gravitational wave strain is defined as
\begin{equation}\label{3.3}
\mathcal{N} = \int_{-\infty}^{t}dt' \psi_4
\end{equation}
The gravitational wave memory contribution to $h_{+}$ polarization can be computed from this bondi news, by integrating the absolute value of bondi news as shown in equation \ref{eq:3.2}. The memory contribution evaluated using Eq.\ref{eq:3.2} will contain contribution from numerical noise, which is unphysical and must be removed from the data. We apply the method described in the following section to remove the unphysical part of GW memory. 
\subsection{Integration of finite length signals in frequency domain}
The unphysical part of mermory essentially lies in lower frequency. A method to remove this unphysical part is given in ref \cite{Reisswig_Pollney2011}. If we remove this low frequency part from data such that the physical memory is still preserved, we can get rid of memory contribution due to numerical noise. 
Consider the Fourier transform, $\mathcal{F}$, applied to an absolutely integrable function $f(t)$,
\begin{equation}
	\tilde{f}(\omega) = \mathcal{F}[f]=\int_{-\infty}^{\infty}\e^{-\iota \omega t}f(t)dt.	
\end{equation} 
The Fourier transform of the time integral is given by
\begin{equation}
	\mathcal{F}\left[\int_{-\infty}^{t}dt' f(t')\right] |_\omega = -\iota \frac{\tilde{f}(\omega)}{\omega}
\end{equation}
The inverse fourier transform of above expression is given by
\begin{equation}\label{3.6}
	\int_{-\infty}^{t}dt'f(t')=\mathcal{F}^{-1}\left[-\mathit{i}\frac{\tilde{f}(\omega)}{\omega}\right] = -\frac{\mathit{i}}{2\pi}\int_{-\infty}^{\infty}\frac{1}{\omega}e^{\mathit{i\omega t}}\tilde{f}(\omega)d\omega
\end{equation}
In the frequency-domain, time integration becomes a simple division by the frequency. Above method is particularly susceptible to low-frequency error, under-resolved high-frequency modes can be aliased into low-frequency modes of the signal. We try to remove these unphysical frequencies by the method described below   
\subsubsection{Fixed frequency integration ($\mathit{FFI}$)}\label{FFI}
The effect of the spurious low-frequency modes caused by spectral leakage or aliasing effects, can be significantly suppressed by use of signal filter. The time integral can be performed in frequency domain as shown in equation \ref{3.6}. The low frequencies which cause the unphysical nonlinear drift can be removed by choosing a cutoff frequency while taking a Fourier transform.

@@

\subsection{Memory calculation from SXS Rumerical Relativity data}
The publicly available SXS catalog provides data for Newman-Penrose scalar component $\psi_4$, decomposed into spin weighted spherical harmonics, extrapolated to infinite radius. We used this data and  performed a frequency domain time integration as described in equation \ref{3.6}, while removing the unphysical frequency by choosing a lower cutoff as show in section \ref{FFI}. The cutoff frequency is chosen as $\omega_0/10$ where $\omega_0$ is the minimum instantaneous GW angular frequency. We first begin by separating the complex Newman-Perose scalar $\psi_4$ data into phase and amplitude part. We compute instantaneous GW angular frequency from this by taking the first time derivative of phase corresponding to $\psi_4$. The instantaneous GW frequency often looks noisy at the beginning of the simulations due to numerical issues. We remove this noise data by choosing a lower and upper cutoff time. Once we find the upper and lower cutoff time, we can apply a frequency domain time integration for $\psi_4$ by choosing a cutoff frequency. After filtering the noise creating frequency we can use equation \ref{eq:3.2} to compute the memory contribution to "plus" polarization of GW signal. The SXS catalog gives data for different parameter values like mass ratio, dimensionless spin parameter etc. The Christodoulou memory computed from NR data starts from zeros using equation \ref{eq:3.2} due to finite simulation time. Chrostodoulou memory is hereditary, hence it's entire past should be included. This can be done by combining the PN expression during early inspiral and NR results for merger part. The method used to connect the memory accumulated during entire life time of BBH coalesce is described in the next section. 
\subsection{Compleate Christodoulou memory during BBH coalescence} 
The PN waveform for Christodoulou are given as a function of parameter $x$ which is related to time by equation \ref{eq:2.14}. This equation can be integrated to obtain a relation between time $t$ and parameter $x$. The Christodoulou memory reaches it's peak value at the merger. We choose the values of integration constants such that the peak in $h_+^{(mem)}$ is reached at time $t=0$. We use our PN results for computing the memory during early inspiral and attach them to NR results near the late inspiral. We compare this memory for different mass ratio and different values of dimensionless spin parameter. Figure \ref{fig:differentmassratio} show the Christodoulou memory for different mass ratios, which shows that the Christodoulou tends to decrease with higher mass ration. The effect of dimensionless spin is shown in figure \ref{fig:differntchivalues}. It can be seen that the spin magnitude has significant effect on nonlinear memory, higher spin magnitude, in case of spin aligned with orbital angular momentum tends to increase the memory contribution. When the spins are anti aligned with orbital angular momentum, it tend to decrease the Christodoulou memory. These results show that it is important to consider the spin contribution when studying the Christodoulou memory in gravitational wave signals.

>>>>

\stm{note everywhere I have added ``the'', that ``upto'' should be ``up to'', etc}
\begin{abstract}
	The Christodoulou memory, which is a nonlinear memory effect sourced by the oscillatory component of the gravitational wave stress tensor, produces a growing, nonoscillatory change in the gravitational wave strain. This results in the permanent displacement of a pair of freely falling test masses after the gravitational wave has passed. While the memory contribution during the early inspiral is well described by the post-Newtonian (PN) approximation, the contribution at merger must be obtained using numerical relativity simulations. The corrections up to third PN order (i.~e.~3PN) for the gravitational wave memory from pairs of nonspinning black holes on quasicircular orbits has been computed by Favata \cite{Favata2009}. Here we include the contribution from the individual black-hole spins at leading order which appears at 1.5PN \stm{what is leading order here, if we then say the memory is 3PN?} and calculate the memory effect up to 3PN order. We also compute the memory contribution during
the merger by using the methods presented in \cite{Favata2009} to calculate the memory content using the oscillatory radiation as input, and
applying that method using publicly available numerical relativity simulations of spinning binaries.
\end{abstract}

@@

where $G^{2-20}_{l'm'lm'-m''-m}$ is the angular integral of equation \ref{eq:2.9} and is given in Appendix \ref{Appendix A}. The above equation can be used to compute the first time derivative of mass-multipole moment by substituting the GW modes $h_{lm}$. When evaluating equation \ref{eq:2.10}, we are only interested in $m=0$ modes. The $m\neq0$ terms yield oscillatory contribution to the waveform polarizations that enter at higher PN orders then the nonoscillatory, $m=0$ terms. The non-vanishing, non-oscillatory modes are $U^{(mem)}_{l0}$ with l-even. These $U^{(mem)}_{l0}$ consist of time-integral of polynomials in $x\equiv(M\omega)^{2/3}$ (where $\omega$ is the orbital angular frequency), with each terms of the form
\begin{equation}\label{eq:2.11}
\int_{-\infty}^{T_{R}} x^{n}dt=\int_{-\infty}^{T_{R}}\frac{x^{n}}{\dot{x}}dx
\end{equation}
After performing the change of variable indicated in above equation \ref{eq:2.11}, evaluation of the resulting integrals requires a model for the frequency evolution of the binary. The adiabatic evolution of the frequency (or $x$) is easily derived from energy balance ($\mathcal{L}_{GW}=-\dot{E}$). and the relation $\dot{x}=-\mathcal{L}_{GW}/(dE/dx)$. The 3.5PN orbital energy and flux are given in Appendix C of reference \cite{Arun2009}. Here we show only up to 1.5 PN terms, the complete expression are show in Appendix \ref{Appendix:C1}. The 1.5 PN expression for energy is given by 
\begin{widetext}
	\begin{align}\label{eq:E}
	E&=-\frac{\eta \, M \, x}{2}\left\{1 + x \left[-\frac{3}{4}-\frac{\eta}{12}\right] + x^{3/2}\left[\left(\frac{8}{3} - \frac{4 \, \eta}{3}\right)\chisdL + \frac{8}{3}\delta\chiadL\right]\right\} \nonumber \\
	\end{align}

@@

\begin{align}\label{eq:2.14}
\frac{dx}{dt} & = \frac{64\, x^5  \, \eta}{5} \left\{1 + x \left[-\frac{743}{336}-\frac{11 \eta }{4} \right] 
+ x^{3/2} \left[4 \pi-\frac{113}{12}\delta \, \chiadL- \left(-\frac{113}{19} + \frac{19 \, \eta}{3}  \right)\chisdL \right] \right\} \nonumber \\ 
\end{align}
\end{widetext}
The results up to 3.5 PN are shows in Appendix \ref{Appendix:C}.
The equation \ref{eq:2.14} can be used to change the integration with respect to time into integration with respect to $x$ as shown in equation in \ref{eq:2.11}. The expressions for $h_{lm}$ whose first time derivatives appear equation \ref{eq:2.10} can be obtained from Ref\cite{Arun2009}. The GW modes for precessing binaries with small inclination angle on nearly circular orbits through 1.5 PN order are given in section IV of ref \cite{Arun2009}. Setting $\iota = 0, \alpha = \pi$ in equation 4.16 and 4.17 gives the result for non-precessing case. In our case the spins are aligned along $z$ axis, so we set $\chi_a^x=\chi_s^x=\chi_a^y=\chi_s^y=0$. We can now write the expression as 
\begin{equation}
h_{lm} = (-1)^{m+1}\frac{(2M\eta x)}{D_L}\sqrt{\frac{16\pi}{5}}e^{\mathit{i}\,m\,\psi}\hat{h}_{lm},
\end{equation}
and $\hat{h}_{lm}$ modes are given by 
\begin{subequations}
	\begin{align}\label{eq:2.15}
	\hat{h}_{22} &= 1 + x\left(-\frac{107}{42} + \frac{55\eta}{42}\right) + x^{3/2}\left(2\pi - \frac{4\,\delta\,\chi^{z}_{a}}{3} + \frac{4}{3}\left(-1+\eta\right)\chi^z_s\right)\\
	\hat{h}_{21} &= \frac{\mathit{i}\delta}{3}\left[x^{1/2} - x\frac{3}{2\delta}\left(\chi^{z}_{a} + \delta\chi^{z}_{s}\right) + x^{3/2}\left(-\frac{17}{28} + \frac{5\eta}{7}\right)\right]\\
	\hat{h}_{32} &= \frac{9}{8}\sqrt{\frac{5}{7}}\left[\frac{8}{27}x\left(1-3\nu\right) + \frac{32}{27}x^{3/2}\eta\,\chi^z_s\right]
	\end{align}
\end{subequations}
The remaining $h_{lm}$ expressions does not contain spin terms and the complete expressions up to 3PN terms can be directly taken from Ref.\cite{Blanchet_Iyer2008}
 We show the $h_{lm}$ expressions for positive $m$ values, but the results for negative $m$ values can be obtained by the relation, $h_{l, -m} = (-1)^l \bar{h}_{lm}$. This results is not in general true for precessing binaries, but it continues to hold true of non-precessing case. The time derivative of $h_{lm}$ modes can be obtained by following the procedure shown in Appendix \ref{Appendix B}. With these modes we use equation \ref{eq:2.10} to evaluate the first time derivative of the memory mass-multipole moment, which is further integrated using equation \ref{eq:2.11}. The results after integration are show in next section.

@@

The above equation requires a time integration of absolute value of $\dot{h}_{22}$, but this is not typically the quantity which is directly computed in numerical simulation. The output of NR simulation is usually in the form of component of curvature tensor, or Zerilli-Moncrief-type variables defined relative to a background. 
\subsection{Evaluating Bondi News from numerical data}
The Bondi News which also the first derivative of gravitational wave strain is defined as
\begin{equation}\label{3.3}
\mathcal{N} = \int_{-\infty}^{t}dt' \psi_4
\end{equation}
The gravitational wave memory contribution to $h_{+}$ polarization can be computed from this bondi news, by integrating the absolute value of bondi news as shown in equation \ref{eq:3.2}. The memory contribution evaluated using Eq.\ref{eq:3.2} will contain contribution from numerical noise, which is unphysical and must be removed from the data. We apply the method described in the following section to remove the unphysical part of GW memory. 
\subsection{Integration of finite length signals in frequency domain}
The unphysical part of mermory essentially lies in lower frequency. A method to remove this unphysical part is given in ref \cite{Reisswig_Pollney2011}. If we remove this low frequency part from the data such that the physical memory is still preserved, we can get rid of memory contribution due to numerical noise. 
Consider the Fourier transform, $\mathcal{F}$, applied to an absolutely integrable function $f(t)$,
\begin{equation}
	\tilde{f}(\omega) = \mathcal{F}[f]=\int_{-\infty}^{\infty}\e^{-\iota \omega t}f(t)dt.	
\end{equation} 
The Fourier transform of the time integral is given by
\begin{equation}
	\mathcal{F}\left[\int_{-\infty}^{t}dt' f(t')\right] |_\omega = -\iota \frac{\tilde{f}(\omega)}{\omega}
\end{equation}
The inverse fourier transform of above expression is given by
\begin{equation}\label{3.6}
	\int_{-\infty}^{t}dt'f(t')=\mathcal{F}^{-1}\left[-\mathit{i}\frac{\tilde{f}(\omega)}{\omega}\right] = -\frac{\mathit{i}}{2\pi}\int_{-\infty}^{\infty}\frac{1}{\omega}e^{\mathit{i\omega t}}\tilde{f}(\omega)d\omega
\end{equation}
In the frequency-domain, time integration becomes a simple division by the frequency. Above method is particularly susceptible to low-frequency error, under-resolved high-frequency modes can be aliased into low-frequency modes of the signal. We try to remove these unphysical frequencies by the method described below   
\subsubsection{Fixed frequency integration ($\mathit{FFI}$)}\label{FFI}
The effect of the spurious low-frequency modes caused by spectral leakage or aliasing effects, can be significantly suppressed by use of signal filter. The time integral can be performed in frequency domain as shown in equation \ref{3.6}. The low frequencies which cause the unphysical nonlinear drift can be removed by choosing a cutoff frequency while taking a Fourier transform.

@@

\subsection{Memory calculation from SXS Rumerical Relativity data}
The publicly available SXS catalog provides data for Newman-Penrose scalar component $\psi_4$, decomposed into spin weighted spherical harmonics, extrapolated to infinite radius. We used this data and  performed a frequency domain time integration as described in equation \ref{3.6}, while removing the unphysical frequency by choosing a lower cutoff as show in section \ref{FFI}. The cutoff frequency is chosen as $\omega_0/10$ where $\omega_0$ is the minimum instantaneous GW angular frequency. We first begin by separating the complex Newman-Perose scalar $\psi_4$ data into phase and amplitude part. We compute instantaneous GW angular frequency from this by taking the first time derivative of phase corresponding to $\psi_4$. The instantaneous GW frequency often looks noisy at the beginning of the simulations due to numerical issues. We remove this noise data by choosing a lower and upper cutoff time. Once we find the upper and lower cutoff time, we can apply a frequency domain time integration for $\psi_4$ by choosing a cutoff frequency. After filtering the noise creating frequency we can use equation \ref{eq:3.2} to compute the memory contribution to "plus" polarization of GW signal. The SXS catalog gives data for different parameter values like mass ratio, dimensionless spin parameter etc. The Christodoulou memory computed from NR data starts from zeros using equation \ref{eq:3.2} due to finite simulation time. Chrostodoulou memory is hereditary, hence it's entire past should be included. This can be done by combining the PN expression during early inspiral and NR results for merger part. The method used to connect the memory accumulated during entire life time of BBH coalesce is described in the next section. 
\subsection{Compleate Christodoulou memory during BBH coalescence} 
The PN waveform for Christodoulou are given as a function of parameter $x$ which is related to time by equation \ref{eq:2.14}. This equation can be integrated to obtain a relation between time $t$ and parameter $x$. The Christodoulou memory reaches it's peak value at the merger. We choose the values of integration constants such that the peak in $h_+^{(mem)}$ is reached at time $t=0$. We use our PN results for computing the memory during early inspiral and attach them to NR results near the late inspiral. We compare this memory for different mass ratio and different values of dimensionless spin parameter. Figure \ref{fig:differentmassratio} show the Christodoulou memory for different mass ratios, which shows that the Christodoulou tends to decrease with higher mass ratios. The effect of dimensionless spin is shown in figure \ref{fig:differntchivalues}. It can be seen that the spin magnitude has significant effect on nonlinear memory, higher spin magnitude, in case of spin aligned with orbital angular momentum tends to increase the memory contribution. When the spins are anti aligned with orbital angular momentum, it tend to decrease the Christodoulou memory. These results show that it is important to consider the spin contribution when studying the Christodoulou memory in gravitational wave signals.
